Covariate analysis of ROSMAP RNASeq data processed using Picard-STAR-HTSeq pipeline from synapse BAM deposited by RUSH

### 2016-11-22 code/Rmd/ROSMAP_RNASeq_Realigned_SynBAM_StarHTSeq.Rmd
Minor modifications

boxplot(VOOM.GENE_EXPRESSION$E, main = 'Distribution of expression')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T, center = T)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=cogdx, shape=Batch, size=age_death))
p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
p
```
Tree based classification of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,FactorCovariates])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix
```{r iterative.norm}
# Primary variable of interest
postAdjustCovars = 'cogdx'

@@

### Residual calculation
Calculate weighted residuals and add back "cogdx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("cogdx", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=cogdx, shape = Batch, size=RINcontinuous))
p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
p
```

@@

# Define biomart object
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")

@@

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
p = p + facet_grid(.~Comparison, scales = 'fixed')
p
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "ROSMAP RNASeq RealignedFromSynBAM PicardStarHTSeq", parentId = parentId)
CODE <- synStore(CODE)

>>>>

for (i in 0:8){
  expr2 = VOOM.GENE_EXPRESSION$E[,rownames(COVARIATES)[(COVARIATES$Batch %in% c(i))]]
  cr = cor(t(expr2))
  hist(cr, main = paste('Batch',i))
}
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T, center = T)

@@

colours <- RColorBrewer::brewer.pal(9, 'Set1')
p <- ggplot(plotdata, aes(x=PC1, y=PC2, shape=cogdx, color = Batch))
p <- p + geom_point(aes(size=RINcontinuous))
p <- p + scale_color_manual(values=colours)
p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
p
```
Tree based classification of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,FactorCovariates])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=12, fig.height=8}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix
```{r iterative.norm}
# Primary variable of interest
postAdjustCovars = c('cogdx', 'Batch')

@@

### Residual calculation
Calculate weighted residuals and add back "cogdx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("cogdx", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')

for (i in 0:8){
  expr2 = RESIDUAL.GENE_EXPRESSION[,rownames(COVARIATES)[(COVARIATES$Batch %in% c(i))]]
  cr = cor(t(expr2))
  hist(cr, main = paste('Batch',i))
}
```

@@

colours <- RColorBrewer::brewer.pal(9, 'Set1')
p <- ggplot(plotdata, aes(x=PC1, y=PC2, shape=cogdx, color = Batch))
p <- p + geom_point(aes(size=RINcontinuous))
p <- p + scale_color_manual(values=colours)
p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
p
```

@@

# Define biomart object
mart <- useMart(host = "jul2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl", biomart = "ENSEMBL_MART_ENSEMBL")

@@

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
p = p + facet_grid(.~Comparison, scales = 'fixed')
p
```

```{r residual.analysis}
expr = RESIDUAL.GENE_EXPRESSION[!duplicated(RESIDUAL.GENE_EXPRESSION$Gene.ID),-c(1:3)]
rownames(expr) = RESIDUAL.GENE_EXPRESSION$Gene.ID[!(duplicated(RESIDUAL.GENE_EXPRESSION$Gene.ID))]

PC = prcomp(expr, scale. = T, center = T)

# Plot first 5 PCs
plotdata <- rownameToFirstColumn(PC$rotation, 'ID') %>% 
    left_join(rownameToFirstColumn(COVARIATES, 'ID'))

barplot((PC$sdev^2/sum(PC$sdev^2))[1:10], main = 'Fraction variance explained (top 10 PC)')
p <- GGally::ggpairs(plotdata, mapping = aes(shape=cogdx, color = Batch, size=RINcontinuous),
                     columns = 2:5, diag = list('continuous' = 'densityDiag'),
                     title = 'First 4 PCs of residuals')
p

lb = median(PC$rotation[,1]) - 2*sd(PC$rotation[,1])
ub = median(PC$rotation[,1]) + 2*sd(PC$rotation[,1])

ind = c(which(PC$rotation[,1] <= lb), which(PC$rotation[,1] >= ub))
writeLines('Samples deviating 2 SD from median of PC1 are')
writeLines(paste(rownames(PC$rotation)[ind], collapse = ', '))

writeLines('Distribution of batches of deviating samples')
kable(rownameToFirstColumn(summary(COVARIATES$Batch[ind]), 'Batch Number'))
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "ROSMAP RNASeq RealignedFromSynBAM PicardStarHTSeq", parentId = parentId)
CODE <- synStore(CODE)
